No one-size-fits-all solution to clean GBIF

### 2020-06-14 manuscript_peerj.Rmd
manuscript resubmission

    @GBIForg2019; @GBIForg2019a; @GBIForg2019b; @GBIForg2019c; @GBIForg2019d; @GBIForg2019e; @GBIForg2019f; @GBIForg2019g; @GBIForg2019h; @GBIForg2019i; @GBIForg2019j; @GBIForg2019k; @GBIForg2019l; @GBIForg2019m; @GBIForg2019n; @GBIForg2019o;@GBIForg2019p;@GBIForg2020, @GBIForg2020a

@@

## Similarities and differences among taxa {-}
The number of records flagged by individual tests was similar across study groups, suggesting that similar problems might be relevant for collections of plants and animals. Therefore, the same filters can be used across taxonomic groups. Some notable exceptions stress the need to adapt each filter to the taxonomic study group to balance data quality and data availability. The high fraction of records filtered by the "basis of record" filter for *Tityus*, Dipsadidae, *Prosthechea* and *Tillandsia*, were mostly caused by a high number of records in these groups based on unknown collection methods, which might be caused by the contribution of specific datasets lacking this information for these groups. The high fraction of records flagged by the "collection year" filter for Dispadidae was caused by a high collection effort in the late 1880s and early 1900s, as can be expected for a charismatic group of reptiles, but also by 500 records dated to the year 1700. The latter records likely represent a data entry error: they are all contributed to GBIF from the same institution, and the institution's code for unavailable collection dates is 1700-01-01 - 2014-01-01, which has likely erroneously been converted to 1700. The high number of species flagged at capital coordinates and within urban areas for the plant groups Iridaceae and *Oocephalus* might be related to horticulture, since at least some species in those groups are commonly cultivated as ornamentals.

@@

# Acknowledgements {-}
We thank GBIF and all data collectors and contributors for their excellent work. We tahnk Town Peterson, Roderic Page and one anonymous reviewer for the helpful comments on an earlier version of this mansucript. This study enrolled participants of the workshop "Biodiversity data: from field to yield" led by Alice Calvente, Fernanda Carvalho, Alexander Zizka, and Alexandre Antonelli through the Programa de Pós Graduação em Sistemática e Evolução of the Universidade Federal do Rio Grande do Norte (UFRN) and promoted by the 6th Conference on Comparative Biology of Monocotyledons - Monocots VI. We thank the Pró-reitoria de Pesquisa and the Pró-reitoria de Pós-graduação of UFRN for financial support (edital 02/2016 - internacionalização). AZ is funded by iDiv via the German Research Foundation (DFG FZT 118), specifically through sDiv, the Synthesis Centre of iDiv. AA is supported by the Swedish Research Council, the Knut and Alice Wallenberg Foundation, the Swedish Foundation for Strategic Research and the Royal Botanic Gardens, Kew. FS was financed by the Coordenação de Aperfeiçoamento de Pessoal de Nível Superior - Brasil (CAPES) - Finance Code 001 and Fundação de Amparo  à Pesquisa do estado de São Paulo (FAPESP) (FAPESP, process 2015/20215-7).

@@

tab1 <- tab1 %>% 
  mutate(taxon =  cell_spec(taxon, "latex", italic = ifelse(.$taxon %in% gens, TRUE, FALSE))) %>% 
  dplyr::select(-contains("pp")) %>% 
  mutate(phylopic = sprintf('\\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/%s.png}', phylopic)) %>% 
  mutate(phylopic = ifelse(grepl("harengula", phylopic), gsub("scale=0.25", "scale=0.18", phylopic), phylopic)) %>%  
  mutate(phylopic = ifelse(grepl("neanuridae", phylopic), gsub("scale=0.25", "scale=0.2", phylopic), phylopic)) %>% 
  mutate(phylopic = ifelse(grepl("tocoyena", phylopic), gsub("scale=0.25", "scale=0.2", phylopic), phylopic)) %>% 
  dplyr::select(phylopic, everything()) %>% 
  select(-order)

@@

kable(tab1, booktabs = T, escape = F, format="latex",
      col.names = linebreak(c("",
                              "Taxon",
                              "Taxon rank",
                              "Realm",
                              "Common name",
                              "'Phylum'",
#                              "Order",
                              "Family")),
        caption = "The study groups and their taxonomy. This study includes three marine and 15 terrestrial taxa, seven of them animals, one group of fungi and ten plants, belonging to 16 different orders. The outlines illustrate the broad taxonomic group. Pictures from www.phylopic.org. Phylopic credits: Diogenidae (Ekaterina Kopeykina (vectorized by T. Michael Keesey), Creative Commons Attribution-NonCommercial-ShareAlike 3.0; Entomobryidae and Neanuridae: Birgit Lang, Creative Commons Attribution 3.0; Tityus: Gareth Monger, Creative Commons Attribution 3.0; \\textit{Gaylussacia}, \\textit{Lepismium}, \\textit{Pilosocereus}: T. Michael Keesey (vectorization) and Nadiatalent (photography), Creative Commons Attribution-ShareAlike 3.0; \\textit{Oocephalus}: Alexander Schmidt-Lebhuhn, Creative Commons Attribution-NonCommercial-ShareAlike 3.0; \\textit{Tillandsia}: Maija Karala, Creative Commons Attribution-NonCommercial-ShareAlike 3.0; all others in the public domain.")%>%
  #landscape()%>%
  kable_styling(font_size = 11) %>%
   row_spec(7, hline_after = TRUE) %>%
   row_spec(8, hline_after = TRUE) %>%
   column_spec(1, width = "1cm", bold = FALSE, italic = FALSE)%>%
   column_spec(2, width = "2.5cm", bold = FALSE, italic = FALSE)%>%
   column_spec(3, width = "1.5cm", bold = FALSE, italic = FALSE)%>%
   column_spec(4, width = "2cm", bold = FALSE, italic = FALSE)%>%
   column_spec(5, width = "2cm", bold = FALSE, italic = FALSE)%>%
   column_spec(6, width = "2cm", bold = FALSE, italic = FALSE)%>%
   column_spec(7, width = "3cm", bold = FALSE, italic = FALSE)
```

>>>>

    @GBIForg2019; @GBIForg2019a; @GBIForg2019b; @GBIForg2019c; @GBIForg2019d; @GBIForg2019e; @GBIForg2019f; @GBIForg2019g; @GBIForg2019h; @GBIForg2019i; @GBIForg2019j; @GBIForg2019k; @GBIForg2019l; @GBIForg2019m; @GBIForg2019n; @GBIForg2019o;@GBIForg2019p;@GBIForg2020, @GBIForg2020a, Goldblatt2008

@@

## Similarities and differences among taxa {-}
The number of records flagged by individual tests was similar across study groups, suggesting that similar problems might be relevant for collections of plants and animals. Therefore, the same filters can be used across taxonomic groups. Some notable exceptions stress the need to adapt each filter to the taxonomic study group to balance data quality and data availability. The high fraction of records filtered by the "basis of record" filter for *Tityus*, Dipsadidae, *Prosthechea* and *Tillandsia*, were mostly caused by a high number of records in these groups based on unknown collection methods, which might be caused by the contribution of specific datasets lacking this information for these groups. The high fraction of records flagged by the "collection year" filter for Dispadidae was caused by a high collection effort in the late 1880s and early 1900s, as can be expected for a charismatic group of reptiles, but also by 500 records dated to the year 1700. The latter records likely represent a data entry error: they are all contributed to GBIF from the same institution, and the institution's code for unavailable collection dates is 1700-01-01 - 2014-01-01, which has likely erroneously been converted to 1700. The high number of species flagged at capital coordinates and within urban areas for the plant groups Iridaceae and *Oocephalus* might be related to horticulture, since at least some species in those groups are commonly cultivated as ornamentals. This was supported by the detailed examination of the data for Iridaceae, which showed that after filtering 1605 records from 69 exotic species remained in the dataset, stressing the importance to address these species in certain taxonomic groups.

@@

# Acknowledgements {-}
We thank GBIF and all data collectors and contributors for their excellent work. We thank Town Peterson, Roderic Page and one anonymous reviewer for the helpful comments on an earlier version of this manuscript. This study enrolled participants of the workshop "Biodiversity data: from field to yield" led by Alice Calvente, Fernanda Carvalho, Alexander Zizka, and Alexandre Antonelli through the Programa de Pós Graduação em Sistemática e Evolução of the Universidade Federal do Rio Grande do Norte (UFRN) and promoted by the 6th Conference on Comparative Biology of Monocotyledons - Monocots VI. We thank the Pró-reitoria de Pesquisa and the Pró-reitoria de Pós-graduação of UFRN for financial support (edital 02/2016 - internacionalização). AZ is funded by iDiv via the German Research Foundation (DFG FZT 118), specifically through sDiv, the Synthesis Centre of iDiv. AA is supported by the Swedish Research Council, the Knut and Alice Wallenberg Foundation, the Swedish Foundation for Strategic Research and the Royal Botanic Gardens, Kew. FS was financed by the Coordenação de Aperfeiçoamento de Pessoal de Nível Superior - Brasil (CAPES) - Finance Code 001 and Fundação de Amparo  à Pesquisa do estado de São Paulo (FAPESP) (FAPESP, process 2015/20215-7).

@@

tab1 <- tab1 %>% 
  mutate(taxon =  cell_spec(taxon, "latex", italic = ifelse(.$taxon %in% gens, TRUE, FALSE))) %>% 
  dplyr::select(-contains("pp")) %>% 
  mutate(phylopic = sprintf('\\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/%s.png}', phylopic)) %>% 
  mutate(phylopic = ifelse(grepl("harengula", phylopic), gsub("scale=0.25", "scale=0.18", phylopic), phylopic)) %>%  
  mutate(phylopic = ifelse(grepl("neanuridae", phylopic), gsub("scale=0.25", "scale=0.2", phylopic), phylopic)) %>% 
  mutate(phylopic = ifelse(grepl("tocoyena", phylopic), gsub("scale=0.25", "scale=0.2", phylopic), phylopic)) %>%
  mutate(phylopic = ifelse(grepl("iridaceae", phylopic), gsub("scale=0.25", "scale=0.075", phylopic), phylopic)) %>% 
  dplyr::select(phylopic, everything()) %>% 
  select(-order)

@@

kable(tab1, booktabs = T, escape = F, format="latex",
      col.names = linebreak(c("",
                              "Taxon",
                              "Taxon rank",
                              "Realm",
                              "Common name",
                              "'Phylum'",
#                              "Order",
                              "Family")),
        caption = "The study groups and their taxonomy. This study includes three marine and 15 terrestrial taxa, seven of them animals, one group of fungi and ten plants, belonging to 16 different orders. The outlines illustrate the broad taxonomic group. Pictures from www.phylopic.org. Phylopic credits: Diogenidae (Ekaterina Kopeykina (vectorized by T. Michael Keesey), Creative Commons Attribution-NonCommercial-ShareAlike 3.0; Entomobryidae and Neanuridae: Birgit Lang, Creative Commons Attribution 3.0; Tityus: Gareth Monger, Creative Commons Attribution 3.0; \\textit{Gaylussacia}, \\textit{Lepismium}, \\textit{Pilosocereus}: T. Michael Keesey (vectorization) and Nadiatalent (photography), Creative Commons Attribution-ShareAlike 3.0; \\textit{Iridaceae} adapted from Goldblatt and Manning (2008); \\textit{Oocephalus}: Alexander Schmidt-Lebhuhn, Creative Commons Attribution-NonCommercial-ShareAlike 3.0; \\textit{Tillandsia}: Maija Karala, Creative Commons Attribution-NonCommercial-ShareAlike 3.0; all others in the public domain.")%>%
  #landscape()%>%
  kable_styling(font_size = 11) %>%
   row_spec(7, hline_after = TRUE) %>%
   row_spec(8, hline_after = TRUE) %>%
   column_spec(1, width = "1cm", bold = FALSE, italic = FALSE)%>%
   column_spec(2, width = "2.5cm", bold = FALSE, italic = FALSE)%>%
   column_spec(3, width = "1.5cm", bold = FALSE, italic = FALSE)%>%
   column_spec(4, width = "2cm", bold = FALSE, italic = FALSE)%>%
   column_spec(5, width = "2cm", bold = FALSE, italic = FALSE)%>%
   column_spec(6, width = "2cm", bold = FALSE, italic = FALSE)%>%
   column_spec(7, width = "3cm", bold = FALSE, italic = FALSE)
```
